---
title: 'Reproducible Research: Peer Assessment 1'
author: "NiazG"
output: word_document
---

## Loading and preprocessing the data

1. Load the data:
``` {r, echo = TRUE}
data <- read.csv("activity.csv", header=TRUE)
```

2. Show the original data:
```{r showdata, echo=TRUE}
head(data)
```

## What is mean total number of steps taken per day?

1. Subset the original data to calculate the total number of steps
``` {r, echo= TRUE}
steps_bDay <- aggregate(steps~date, data, sum)
head(steps_bDay)
```

2. Make a histogram of the total number of steps taken each day

``` {r, echo = TRUE}
hist(steps_bDay$steps, col = "Darkgreen"
     , xlab = "Steps", main = "Total Number of Steps Per Day")
```

3. Calculate the mean and median number of steps taken each day
``` {r, echo = TRUE}
mean_steps <- mean(steps_bDay$steps)
mean_steps
median_steps <- median(steps_bDay$steps)
median_steps
```

## What is the average daily activity pattern?

1. Subset the original data to calculate the average number of steps taken by interval

``` {r, echo = TRUE}
avg_bDay <- aggregate(steps~interval, data, mean)
head(avg_bDay)
```

2. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

``` {r, echo = TRUE}
with(avg_bDay, plot(interval,steps,col = "Darkblue", type = "l", 
                    xlab = "Interval" ,ylab="Average number of steps", 
                    main = "Time series plot of the average number of steps taken"))
```

3. Finding the  interval that, on average, contains the maximum number of steps
``` {r, echo=TRUE}
int_max_steps <- avg_bDay$interval[max(avg_bDay$steps)]
int_max_steps
```

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA's)
``` {r, echo = TRUE}
na_rows <-  sum(is.na(data$steps))
na_rows
```

2. Make a new column with mean of steps taken each day
``` {r, echo = TRUE}
data$newSteps <- avg_bDay$steps
head(data)
```

3. Fill the NA's in steps
``` {r, echo = TRUE}
data$steps[is.na(data$steps)] <- data$newSteps
head(data)
```

4. Create a new dataset that is equal to the original dataset but with the missing data filled in
``` {r, echo = TRUE}
data$newSteps <- NULL
newdata <- data
```

5. Subset in the new data set to calculate the total number of steps per day
```{r, echo=TRUE}
steps_new <- aggregate(steps~date, newdata, sum)
head(steps_new)
```

6. Make a histogram of the total number of steps taken each day
```{r, echo=TRUE}
hist(steps_new$steps,col = "Darkblue",
     xlab = "Steps", 
     main = "Total Number of Steps Per Day after filling Na's")
```


7. Calculate the differences of the mean and median between the first and second part
```{r, echo = TRUE}
mean_new <- mean(steps_new$steps)
mean_new
median_new <- median(steps_new$steps)
median_new
```

8. Calculate the differences of the mean and median between the first and second part
``` {r, echo = TRUE}
mean_steps - mean_new
median_steps - median_new
```

## Are there differences in activity patterns between weekdays and weekends?
For this part, I do not use the weekdays() function, instead I use the isWeekday() function from the timeDate package.

1. Install and load the timeDate package
install.packages("timeDate")
```{r, echo=TRUE}
library(timeDate)
```

2. Create a new column, and use the isWeekday() function to check if the date is weekday or weekend
```{r, echo=TRUE}
newdata$Weekday <- isWeekday(newdata$date)
head(newdata)
```

3. Subset and calculate the average steps for weekday and weekend

Weekday
```{r, echo=TRUE}
weekday <- subset(newdata, newdata$Weekday == "TRUE")
weekdayMean <- aggregate(steps ~ interval, data = weekday, mean)
head(weekdayMean)
```

Weekend
```{r, echo=TRUE}
weekend <- subset(newdata, newdata$Weekday == "FALSE")
weekendMean <- aggregate(steps ~ interval, data = weekend, mean)
head(weekendMean)
```

4. Make the panel plot to calculate the average number of steps taken for weekday and weekend
```{r, echo=TRUE}
layout(matrix(c(1,1,2,2), 2, 2, byrow = TRUE))
plot(weekdayMean$interval, weekdayMean$steps
     , xlab = "interval", ylab = "Number of steps"
     , main ="Weekday", col ="blue", type="l") 
plot(weekendMean$interval, weekendMean$steps
     , xlab = "interval", ylab = "Number of steps"
     , main ="Weekend", col ="red", type="l")
```


















